Demystifying the Computer Science \\ PhD Admission in the US\\{\large A Handbook for International \emph{and Domestic} Students

### 2024-10-22 demystify.tex
updates

  \item [PI]  Principal Investigator is lead researcher on a grant or research project, e.g., a PI on an NSF grant.  Prospective students often use this term to refer to a professor they are interested in working with.  However, we do not use it this way in academia, e.g., ``who is your PI?'' might only make sense to fellow Reddit applicants, but might not be understood by others, ``who is your adviser?'' is what you want to use.
    \item[STEM] Science, Technology, Engineering, and Mathematics. A term used to group together these academic disciplines. CS is considered a STEM field.
  \item[URM] Underrepresented Minority. Refers to groups that are underrepresented in higher education, e.g., Black, Hispanic, and Native American students.
  \item[NSF] National Science Foundation. A US government agency that supports research and education in science and engineering. NSF is a major source of funding for RAs in CS research. There are other agencies that support RAs but they might require US citizenship or permanent residency.
        \item [R1/R2 Universities] Research 1 (R1) universities are universities with the highest level of research activity. R2 universities have a high level of research activity but not as high as R1 universities.  R1 universities are typically larger, and have more funding and resources for research.  R2 universities are also good but might have fewer resources and funding.

@@

While both MS and PhD programs are graduate degrees, they are \emph{very} different in terms of objective, admission requirements, course requirements, duration, and funding. This section provides guidance on applying to MS programs in CS.

@@

\begin{table}
    \caption{MS vs. PhD}\label{tab:phd-vs-ms}
    \begin{tabular}{c|c|c}
    \toprule
    \textbf{} & \textbf{MS} & \textbf{PhD} \\
    \midrule
    Objective & Industry & Research \\
    Admission Req & Research experience & No research experience \\
    Coursework Req & Yes & Yes (research is much more important)  \\
    Duration & 2 years & 5--7 years \\
    Adviser Req & No & Yes \\
    Funding & No & Yes \\
    \bottomrule
    \end{tabular}
  \end{table}

@@

Except for a very few highly selective programs, MS programs are typically much less competitive than PhD programs. Some refer to MS programs as a \emph{cash cow} for the department, as MS students are often not funded and have to pay tuition (i.e., you're likely to get in if you can afford it).

While admission requirements are similar to PhD programs (e.g., GPA, SOP, LORs), research is not a focus in MS programs. In most cases, the main requirements are that you have sufficient background in CS (e.g., through your undergrad degree).

\subsection{Admission Committee}\label{sec:ms-adcom}
MS admission also involves an adcom that reviews applications and makes admission decisions. However, MS admission is typically rolling (i.e., they review applications as they come in and make decisions throughout the admission cycle).  Moreover, unlike PhD that has multiple reviewers for each application, an MS application typically involves only one reviewer and does not review much time to review compared to a PhD application (\autoref{sec:ievaluate}).

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically not funded through either RA or TA.
\begin{itemize}
\item (RA) Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by the time they are, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.
\item (TA) MS students are typically not given TA positions because PhDs are given priority. Many depts. even have shortage of TA positions for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

@@

This section lists some cultural issues that students, especially international ones, might want to pay attention to. These issues apply in general and not just PhD admission.

@@

  Moreover, avoid asking personal questions, e.g., about salary, age, relationship status, or health, as these are considered private (talking about kids' activities or schools are OK). Also, do not talk about politics or religion in general (in fact, we often do not talk about these with our family and friends to avoid conflicts).  Sometimes foreigners are surprised by how Americans do not talk about their personal lives, e.g., sharing details about their families, health, or relationships, and that their conversations are often not very ``deep'' or ``mind-provoking''. This is just a cultural norm about privacy and personal space.

@@

  \paragraph{Car Dependency} Most of the US is highly car-dependent. If you do not have a car, you will need to rely on friends, Uber, or public transportation, which can be inconvenient and time-consuming. Most international students end up getting a driver's license and buying a car.

	%\paragraph{Individualism} The strong culture of individualism in the US can be both liberating and isolating. For example, Americans are encouraged to express their opinions and pursue their goals, and are more focused on their personal rights (e.g., they do not want people do things in front of their house).

>>>>

  \item [PI]  Principal Investigator is lead researcher on a grant or research project, e.g., a PI on an NSF grant.  Prospective students often use this term to refer to a professor they are interested in working with.  However, we do not use it this way in academia, e.g., ``who is your PI?'' might only make sense to fellow Reddit applicants, but might not be understood by others, ``who is your adviser?'' is what you want to use.
    \item[STEM] Science, Technology, Engineering, and Mathematics. A term used to group together these academic disciplines. CS is considered a STEM field.
  \item[URM] Underrepresented Minority. Refers to groups that are underrepresented in higher education, e.g., Black, Hispanic, and Native American students.
  \item[NSF] National Science Foundation. A US government agency that supports research and education in science and engineering. NSF is a major source of funding for RAs in CS research. There are other agencies that support RAs but they might require US citizenship or permanent residency.

  \item [R1/R2 Universities] Research 1 (R1) universities are universities with the highest level of research activity. R2 universities have a high level of research activity but not as high as R1 universities.  R1 universities are typically larger, and have more funding and resources for research.  R2 universities are also good but might have fewer resources and funding.

@@

While both MS and PhD programs are graduate degrees, they are \emph{very different} in terms of objective, admission requirements, course requirements, duration, and funding. This section provides guidance on applying to MS programs in CS.

@@

\begin{table}
    \caption{MS vs. PhD}\label{tab:phd-vs-ms}
    \centering
    \small
    \begin{tabular}{c|c|c}
    \toprule
    \textbf{} & \textbf{MS} & \textbf{PhD} \\
    \midrule
    Objective & Industry & Research \\
    Admission Req & Research experience & No research experience \\
    Coursework Req & Yes & Yes (research is much more important)  \\
    Duration & 2 years & 5--7 years \\
    Adviser Req & No & Yes \\
    Funding & No & Yes \\
    \bottomrule
    \end{tabular}
  \end{table}

@@

Except for a very few highly selective programs, MS programs are typically much less competitive than PhD programs, i.e., you're likely to get in if you can afford it. Some refer to MS programs as a \emph{cash cow} for the department, as MS students are often not funded and have to pay tuition.

While admission requirements are similar to PhD programs (e.g., GPA, SOP, LORs), research is not a focus in MS programs. In most cases, the main requirements are that you have sufficient background in CS, e.g., through your undergrad degree.

\subsection{Admission Committee}\label{sec:ms-adcom}
MS admission also involves an adcom that reviews applications and makes admission decisions. However, MS admission is typically rolling (i.e., they review applications as they come in and make decisions throughout the admission cycle).  Moreover, unlike PhD that has multiple reviewers for each application, an MS application typically involves only one reviewer and does require much time to review compared to a PhD application (\autoref{sec:ievaluate}).

@@

\subsection{Funding}\label{sec:ms-funding}
Unlike PhD programs, which often have funding (\autoref{sec:funding}), MS students are typically not funded. The reason is because MS students do not focus on research and thus are not funded through RA and PhD students have priority for TA positions. 
\begin{itemize}
\item (RA) Profs. are not willing to take MS students as RAs because they are not around long enough to be productive. It can take a while for a student to get up to speed and start being productive, and by the time they are, they are about to graduate. Moreover, the goal of most MS students is to get a job, not to do research, so they are not as motivated to do research.
\item (TA) MS students are typically not given TA positions because PhDs are given priority. Many depts. even have shortage of TA positions for PhD students and so they cannot afford to give them to MS students.
\end{itemize}

@@

This section lists some general cultural issues that students, especially international ones, might want to pay attention to.

@@

Moreover, avoid asking personal questions, e.g., about salary, age, relationship status, or health, as these are considered private.  Talking about kids' activities or schools are OK. Also, do not talk about politics or religion.  In fact, we often do not talk about these taboo subjects with our family and friends to avoid conflicts.

Sometimes foreigners are surprised by how Americans do not talk about their personal lives, e.g., sharing details about their families, health, or relationships, and that their conversations are often not very ``deep'' or ``mind-provoking''. This is just a cultural norm about privacy and personal space.

@@

  \paragraph{Car Dependency} Most places in the US are highly car-dependent. If you do not have a car, you will need to rely on friends, Uber, or public transportation, which can be inconvenient and time-consuming. Many international students end up getting a driver's license, which is highly convenient and replaces many documents (e.g., ID, passport) and buying a car.

  %\paragraphP{Conversation} Americans are generally friendly and open.  However, there are certain things that we rarely talk about, e.g., salary, relationship status, or health. 
  These are considered private and should not be asked.  Also, do not talk about politics or religion in general (in fact, we do not even talk about these with our family and friends to avoid conflicts).

  %Sometimes foreigners are surprised by how Americans do not talk about their personal lives, e.g., sharing details about their families, health, or relationships, and that their conversations are often not very ``deep'' or ``mind-provoking''. This is just a cultural norm about privacy and personal space.

  %This can be surprising for people from countries where strangers do not typically talk to each other.  Americans are also generally informal and use first names, even with people they have just met.  This can be surprising for people from countries where titles and last names are used, even with close friends.
	%\paragraph{Individualism} The strong culture of individualism in the US can be both liberating and isolating. For example, Americans are encouraged to express their opinions and pursue their goals, and are more focused on their personal rights (e.g., they do not want people do things in front of their house).
